Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-12-03 rnni_convexity.tex
update discussion

\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{RNNI}
\vspace{12pt}
\caption{Two types of operation that define edges of the $\rnni$ graph,
The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the blue edges.}
\label{fig:RNNI}
\end{figure}

@@

Within this paper we established some geometric properties of the $\rnni$ graph.
We furthermore pointed out some differences between this graph on ranked trees and the well studied $\nni$ graph on unranked trees.
This is of particular interest for analysing the complexity of computing distances in $\rnni$ as this problem is known to be $\np$-complete in $\nni$.
Our finding that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}) suggests that there is a difference in complexity between these two tree spaces.
This leaves hope that computing distances in $\rnni$ is, contrary to this problem in $\nni$, not $\np$-complete.
The question of how hard this problem actually is still remains open and will be subject of further research.

Our work in this paper highlights the importance of algorithms for computing paths in $\rnni$ in order to prove geometric properties such as diameter and radius.
One of the algorithms that has been introduced, $\findpath$, computed exact distances for all pairs of trees on up to six taxa.
For future investigations of the $\rnni$ space this algorithm might be a useful tool.

>>>>

\begin{figure}[H]
\centering
\includegraphics[width=0.8\textwidth]{RNNI}
\vspace{12pt}
\caption{Two types of operation that define edges of the $\rnni$ graph.
The rank move swaps the rank of the two highlighted nodes and the $\nni$ moves are performed on the blue edges.}
\label{fig:RNNI}
\end{figure}

@@

The aim of this paper was to gain better understanding of the $\rnni$ space of ranked phylogenetic trees.
Besides proving some geometrical properties of $\rnni$, we also introduce some algorithms in this paper.
Our work highlights the importance of these algorithms as they are important for proving geometric properties such as diameter and radius of the $\rnni$ graph.
The most important algorithm we introduced might be $\findpath$, which computes exact distances for all pairs of trees on up to six taxa.
This algorithm might be a useful tool for future investigations of the $\rnni$ space.

We used the algorithms introduced in Section~\ref{section:algorithms} to prove some geometric properties of the $\rnni$ graph such as diameter and radius.
Moreover, we discovered that the set of caterpillar trees is convex in $\rnni$ (Theorem~\ref{thm:caterpillar_convex}).
With this result and the algorithm $\csort$ for computing shortest caterpillar paths (Section~\ref{section:alg_csort}) we can compute $\rnni$ distances between caterpillar trees in polynomial time.

This fact that computing $\rnni$ distances between caterpillar trees only requires polynomial running time already suggests that there is a difference in complexity of computing distances in $\rnni$ and $\nni$.
In Section~\ref{section:cluster_theorem} we supported this claim further by discussing the proof of $\np$-completeness of computing distances in $\nni$ and showing that this proof does not work for $\rnni$.
Specifically, we claimed the Cluster Theorem (Conjetcure~\ref{conjecture:cluster_theorem}), which does not hold in $\nni$, but does hold in the $\rnni$ for trees on up to six taxa \autocite{Collienne2019}.
Though the Cluster Theorem remains a conjecture, the results of this paper suggest that it is true.

\subsection{Open Problems}

In Section~\ref{section:cluster_theorem} we claim that the Cluster Theorem holds for $\rnni$.
We could show that the previous version of it, the Split Theorem as it was conjectured in \autocite{Gavryushkin2018-ol}, is not true.
If the Cluster Theorem actually is true still remains an open question.
Proving it will be a big step towards finding the complexity of computing $\rnni$ distances.

This question of whether computing $\rnni$ distances is $\np$-hard is another problem for future research.
Though we did not aim to answer this question in this paper, we made some progress towards solving it by establishing some geometric properties of the $\rnni$ graph.
Comparing these results with what is known for the plain $\nni$ graph makes us assume that the complexity of computing shortest path in $\rnni$ might differ from $\nni$.
We also suspect that the fact that computing distances between caterpillar trees can be done in polynomial time might be the key for further investigation of the complexity of computing $\rnni$ distances.
